---
title: "Installation"
subtitle: "Steps to install the package and gointo user interface"
title-block-banner: true
title-block-banner-color: white
title-block-categories: false

page-layout: full
format:
  html: 
    margin-top: 0em
    margin-bottom: 0em
    padding-top: 0em
    padding-bottom: 0em
    minimal: true
    smooth-scroll: true
    fig-responsive: true
    toc-location: right
    toc-depth: 5
    toc-title: Installation
    code-fold: show
    code-summary: "Show the code"
    code-overflow: wrap
    code-tools: true
    code-copy: true
    highlight: tango
    df-print: paged  ## kable is another option How the dataframe looks like 
    standalone: false  ### specifies if all assets and libraries must be integrated into the output html file as a standalone document.
    fig-align: right
    theme:  solar
geometry: margin=lin  
---

### Install packages

Please run the following code to install the packages needed. This needs to be run in your R console.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output

### I may need mix Omics and impute not sure
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("mixOmics")
BiocManager::install("impute")
BiocManager::install("preprocessCore")
BiocManager::install("GO.db")

install.packages("remotes")
remotes::install_gitlab("YingxiaoYan/TriplotGUI")
```

Sometimes it pops up in your console to ask you *which package you would like to update*. We recommened you to chose **3.None**

Sometimes it pops up in your console to ask *Do you want to install from soutce packages which need compliation* . We recommended you to chose **no** 

### Load libraries 

Run the following code in your R console.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output

library(TriplotGUI)
```

### Download the example data for user interface

Go to my [git repository](https://gitlab.com/YingxiaoYan/triplotgui/-/blob/main/Example_data.rar?ref_type=heads){target="_blank"}.
Click the download icon at the right side and download the Exampledata.rar. Unzip it and you will find the data that will be used as example in the tutorial.


### Activate the user interface

Run the following code in your R console.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output

options(shiny.maxRequestSize=100*1024^2)
TriplotGUI_shiny_tryout()

```


