---
title: "Installation"
subtitle: "Steps to install the package and gointo user interface"
title-block-banner: true
title-block-banner-color: white
title-block-categories: false

page-layout: full
format:
  html: 
    margin-top: 0em
    margin-bottom: 0em
    padding-top: 0em
    padding-bottom: 0em
    minimal: true
    smooth-scroll: true
    fig-responsive: true
    toc-location: right
    toc-depth: 5
    toc-title: Installation
    code-fold: show
    code-summary: "Show the code"
    code-overflow: wrap
    code-tools: true
    code-copy: true
    highlight: tango
    df-print: paged  ## kable is another option How the dataframe looks like 
    standalone: false  ### specifies if all assets and libraries must be integrated into the output html file as a standalone document.
    fig-align: right
    theme:  solar
geometry: margin=lin  
---

# Install packages

Please run the following code to install the packages needed. This needs to be run in your R console.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output

### I may need mix Omics and impute not sure
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("mixOmics")
BiocManager::install("impute")
BiocManager::install("preprocessCore")
BiocManager::install("GO.db")

install.packages("remotes")
remotes::install_gitlab("YingxiaoYan/TriplotGUI")
```

Things may pop up in your console and ask you for some choices, we provide some recommedations here. We are conservative in the choices to avoid any unforseen conflict and complications in packages.

<ul>

<li>Sometimes it pops up in your console to ask you *which package you would like to update*. We recommened you to chose **3.None**</li>

<li>Sometimes it pops up in your console to ask *Do you want to install from source packages which need compliation* . We recommended you to chose **no**</li>

<li>Sometimes it pops up in your console to ask *Update all/some/none?\[a/s/n\]* . We recommended you to chose **n**</li>

</ul>

# Load libraries

Run the following code in your R console.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output

library(TriplotGUI)
```

# Download the example data for user interface

Go to my [git repository](https://gitlab.com/YingxiaoYan/triplotgui/-/blob/main/Example_data.rar?ref_type=heads){target="_blank"}. Click the download icon at the right side and download the Exampledata.rar. Unzip it and you will find the data that will be used as example in the tutorial.

# Change the maximize size of data file allowed

The default of our package has allows data of maximize size of 5 MB. To change that, you could specify the maximize file size you prefer using the code below. Run your code in the console. However, we do not recommend to use data with more than 5 MB in our tool. The bigger the dataset, the slower it will be. It might be beneficial for you to do some variable selection before putting the data in out tool.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output


filesize=100  ##MB
options(shiny.maxRequestSize=filesize*1024^2)

```

# Activate the user interface

Run the following code in your R console.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output


TriplotGUI_shiny_tryout()

```
