---
title: "Installation"
filters:
   - lightbox
lightbox: auto
subtitle: "Install TriplotGUI as a stand-alone application or use it on webserver"
title-block-banner: true
title-block-banner-color: white
title-block-categories: false

page-layout: full
format:
  html: 
    margin-top: 0em
    margin-bottom: 0em
    padding-top: 0em
    padding-bottom: 0em
    minimal: true
    smooth-scroll: true
    fig-responsive: true
    toc-location: right
    toc-depth: 5
    toc-title: Installation
    code-fold: show
    code-summary: "Show the code"
    code-overflow: wrap
    code-tools: true
    code-copy: true
    highlight: tango
    df-print: paged  ## kable is another option How the dataframe looks like 
    standalone: false  ### specifies if all assets and libraries must be integrated into the output html file as a standalone document.
    fig-align: right
    theme:  solar
geometry: margin=lin  
---

The TriplotGUI tool can be accessible online at [here](https://yingxiao.shinyapps.io/triplotgui/){target="_blank"}. If you want to use TriplotGUI as a stand-alone application (using it only on a local computer), please follow the procedure below.

# Install packages

Please run the following code to install the packages needed. This needs to be run in your R console.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output

install.packages("remotes")
remotes::install_gitlab("YingxiaoYan/TriplotGUI")
```

Things may pop up in your console and ask you for some choices, we provide some recommedations here. We are conservative in the choices to avoid any unforseen conflict and complications in packages.

<ul>

<li>Sometimes it pops up in your console that asks you *which package you would like to update*. We recommened you to chose **3.None**</li>

<li>Sometimes it pops up in your console that asks *Do you want to install from source packages which need compliation* . We recommended you to chose **no**</li>

<li>Sometimes it pops up in your console that asks *Update all/some/none?\[a/s/n\]* . We recommended you to chose **n**</li>

</ul>

# Load libraries

After installing the package, run the following code in your R console.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output

library(TriplotGUI)
```



# Change the maximize size of data file allowed

The default of our package has allows data of maximize size of 5 MB. To change that, you could specify the maximize file size you prefer using the code below. Run your code in the console. However, we do not recommend to use data with more than 5 MB in our tool. The bigger the dataset, the slower it will be. It might be beneficial for you to do some variable selection before putting the data in out tool.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output

filesize=100  ##MB
options(shiny.maxRequestSize=filesize*1024^2)

```

# Activate the user interface

To activate the graphical user interface, run the following code in your R console.

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#| classes: styled-output


TriplotGUI_shiny()

```
