---
title: "Publication & honor"
filters:
   - lightbox
lightbox: auto
subtitle: "Publications and honors"
title-block-banner: true
title-block-banner-color: white
comments: false
page-layout: full
format:
  html: 
    page-layout: full
    margin-left: 0em
    margin-right: 0em
    margin-top: 0em
    margin-bottom: 0em
    padding-top: 0em
    padding-bottom: 0em
    minimal: true
    smooth-scroll: true
    fig-responsive: true
    toc-location: left
    toc-depth: 5
    toc-title: Awesome me
    number-sections: true  ## add number in the session or not
    number-depth: 4
    code-fold: true
    code-summary: "Show the code"
    code-overflow: wrap
    code-tools: true
    code-copy: true
    highlight: tango
    df-print: paged  ## kable is another option How the dataframe looks like 
    standalone: false  ### specifies if all assets and libraries must be integrated into the output html file as a standalone document.
    fig-align: right
    theme:  solar
geometry: margin=lin  
#execute:
#  freeze: true  # never re-render during project render
---

## Publications

Luyan Wu, Wenhui Zeng, Liandong Feng, Yuxuan Hu, Yidan Sun, <strong>Yingxiao Yan</strong>, HongYuan Chen & Deju Ye. <a href="https://onlinelibrary.wiley.com/doi/10.1111/joim.13540">An activatable radiometric near-inflared fluorescent probe for hydrogen sulfide imaging in vivo.</a> Science China Chemistry. 2020;63(5):741-750 <br /> <br /> <strong>Yan Y</strong>, Smith E, Melander O, Ottosson F. <a href="https://link.springer.com/article/10.1007/s11426-019-9689-4#citeas">The association between plasma metabolites and future risk of all‐cause mortality.</a> Journal of Internal Medicine. 2022;292(5):804-815. <br /> <br /> Schillemans T, <strong>Yan Y</strong>, Ribbenstedt A, Donat-Vargas C, Lindh C, Kivirante H, Wolk A, et al.OMICs signatures linking persistent organic pollutants to cardiovascular disease in the Swedish Mammography Cohort \[Submitted\]

## Conference and awards

<ul>

<li><a href="https://www.metabolomics2022.org/">Metabolomics 2022</a>, Valencia, Spain. <em>- Speaker, with award</em></li>

<li><a href="https://www.sbw2022.com/">Swedish Bioinformatics workshop 2022</a>,Umeå, Sweden. <em>- Speaker</em></li>

<li>Nordic rye forum 2022, Stockholm, Sweden <em>- Speaker</em></li>

</ul>

## Key competence

<ul>

<li>R/shiny (data cleaning, manipulation, analysis, and visualization) <em>- Expert</em></li>

<li>Biostatistics and epidemiological research <em>- Expert</em></li>

<li>SPSS, Excel & Microsoft Office suite <em>- Expert</em></li>

<li>python <em>- Intermediate</em></li>

</ul>

## Languages

<ul>

<li>Mandarin (Native speaker)</li>

<li>English (Native level)</li>

<li>Swedish (Intermediate/B1 level, SAS certificated)</li>

</ul>

## References

Filip Ottoson <br />Postdoctoral Researcher – <em>Lund University</em> <br />Email: filip.ottoson\@med.lu.se <br />Tel: +46739699000 <br /> <br />Carl Brunius <br />Associate Professor – <em>Chalmers University of Technology</em> <br />Email: calbrunius\@chalmers.se <br />Tel:

<footer>

<p>© 2023 Yingxiao Yan ★★★★★. All rights reserved</p>

</footer>

::: {#news}
## News

-   **January 23, 2024**: I will be at [Bayes on the Beach 2024](https://research.qut.edu.au/qutcds/bayes-onthe-beach/) in February. I will give a talk about reliable amortized Bayesian inference with neural networks and co-lead a workshop on amortized Bayesian workflows.

-   **December 2023**: We launched the [BayesFlow Forums](https://discuss.bayesflow.org/)! The BayesFlow Forums provide a community for asking and answering questions about all aspects of [BayesFlow](https://www.bayesflow.org) and amortized Bayesian workflows in general. Hop over and join the discussion!

-   **October 11, 2023**: In our recent preprint, we make simulation-based inference more data-efficient by leveraging self-consistency properties of the Bayesian joint model. You find the preprint on [arXiv](https://arxiv.org/abs/2310.04395). Update: Published in the NeurIPS UniReps workshop.

-   **September 22, 2023**: Our paper "Detecting Model Misspecification in Amortized Bayesian Inference with Neural Networks" has been awarded the [DAGM GCPR Honorable Mention](https://x.com/MarvinSchmittML/status/1706281145501933850) at this year's German Conference on Pattern Recognition. Huge thanks to my great co-authors!

-   **September 6, 2023**: Our paper "Detecting Model Misspecification in Amortized Bayesian Inference with Neural Networks" has been accepted with oral presentation at the [German Conference for Pattern Recognition 2023](https://www.dagm-gcpr.de/year/2023). Learn more at the [project website](projects/#sbi-model-misspecification).

-   **June 21, 2023**: I will be at the [ELLIS Doctoral Symposium 2023](https://ellis.eu/events/ellis-doctoral-symposium-2023) in Helsinki from August 28 to September 1, 2023. Absolutely thrilled to meet other PhD students and researchers and talk about machine learning research in beautiful Helsinki!

-   **May 9, 2023**: Our paper "JANA: Jointly Amortized Neural Approximation of Complex Bayesian Models" has been accepted to UAI 2023 (Uncertainty in AI). Check out the [project page](projects/#jana-jointly-amortized-neural-approximation-of-complex-bayesian-models) to find out more!

-   **March 29, 2023**: I'll be presenting our paper on *Meta-Uncertainty in Bayesian Model Comparison* at AISTATS 2023 from April 25-27 in Valencia, Spain. You find links to the paper and code on my <a href="projects/index.qmd#meta-uncertainty-BMC" target="_blank" style="color: #E6AA04;">Projects</a> page or on the dedicated <a href="https://meta-uncertainty.github.io/" target="_blank" style="color: #E6AA04;">paper website</a>. If you see me around at AISTATS, let's chat!
:::

## Featured Blog Posts

::: {#blog-listing}
:::

```{r }
#| eval: false
#| echo: true
#| warning: false
#| error: true
#don't run code, but show code
3
```

```{r}
#| eval: true
#| echo: true
#| output: false
#run code, and show code, don't show output
4
```

```{r}
#| eval: true
#| echo: false
#| output: true
#run code, not show code, but show output
1+1
```

```{r}
#| eval: true
#| echo: true
#| output: true
#run code, and show code, and show output
5
```
